Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-11 MS/box.Rmd
boxes done - a bit of corrections for ms.Rmd too

## *BOX: An example of how to obtain sampling variance for an effect size - let's create one yourself*

$$
\text{ARR} = \frac{ \text{CT}_{\text{max}[T_2]} - \text{CT}_{\text{max}[T_1]}}{T_2 - T_1},
$$
where *T* stands for temperature in Celsius and $T_2 > T_1$.

Let female ARR and male ARR, $\text{ARR}_{\text{f}}$ and $\text{ARR}_{\text{m}}$, respectively and then;

$$
\text{ARRD} = \text{ARR}_{\text{f}} - \text{ARR}_{\text{m}}.
$$
Importantly, ARRD behaves like traditional effect sizes such as standardized mean difference (aka, Cohen's $d$ or Hedges' $g$) and lnRR (log response ratio).

### Sampling variance for ARRD

#### Independent case

We assume all animals measured at at *T~1~* and *T~2~* are different (independent)

$$
s^2(\text{ARRD}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{sd^2_{f[T_1]}} {n_{f[T_1]}} + \frac{sd^2_{f[T_2]}} {n_{f[T_2]}} +
\frac{sd^2_{m[T_1]}} {n_{m[T_1]}} + \frac{s^2_{m[T_2]}} {n_{m[T_2]}}\right),
$$
where $s^2$ is the sampling variance, $sd$ is the sample standard deviation, $n$ is the sample size.

#### Dependent case

We may have cases where the same animals are measured  both at *T~1~* and *T~2~*. So we have $n_f$ and $n_m$, which need to be the same at both temperatures. Then, we need to have an estimate of correlation between measurements of the same individuals ($r_{f[T_1T_2]}$ for females and $r_{m[T_1T_2]}$ for males).

$$
s^2(\text{ARRD}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{sd^2_{f[T_1]} + sd^2_{f[T_2]} - r_{f[T_1T_2]}sd_{f[T_1]}sd_{f[T_2]}} {n_{f}}   +
\frac{sd^2_{m[T_1]} + sd^2_{m[T_2]} - r_{m[T_1T_2]}sd_{m[T_1]}sd_{m[T_2]}} {n_{m}}
\right).
$$

In reality, we often do not know the correlation $r_{[T_1T_2]}$ so we will probably use 0.5.

## AARR (ARR -1)

### Point estimate

$ARRD = 1$ represents perfect acclimation. We want to make 0 this value by substracting 1 and call this AARR (adjusted acclimation response ratio)

$$
\text{AARR} = \text{ARR} -1.
$$

Note AARD like ARRD behaves like tranditional effect statistics.

### Sampling variance for AARR

#### Independent case

$$
s^2(\text{AARR}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{sd^2_{[T_1]}} {n_{[T_1]}} + \frac{sd^2_{[T_2]}} {n_{[T_2]}} \right).
$$

#### Dependent case

$$
s^2(\text{AARR}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{sd^2_{[T_1]} + sd^2_{[T_2]} - r_{[T_1T_2]}sd_{[T_1]}sd_{[T_2]}} {n} 
\right).
$$

##### Basics about variance

See https://en.wikipedia.org/wiki/Variance

Adding a constant ($a$) does not change the variance of a random varaible $X$.
$$
\operatorname{Var}(X+a)=\operatorname{Var}(X)
$$

Multipling a constant ($a$) does increase the variance of $X$ by $a^2$.

$$
\operatorname{Var}(aX)=a^2\operatorname{Var}(X)
$$

When we consider two random variables $X$ and $Y$, we need to know their covariance.

$$
\operatorname{Var}(aX\pm bY)=a^2\operatorname{Var}(X)+b^2\operatorname{Var}(Y) \pm 2ab\operatorname{Cov}(X,Y)
$$
## Note

1. General Slope - some basics
2. ARR
3. Differences between slopes
4. ARRD
5. Delta method

>>>>

## *BOX: Example of how to derive sampling variance*

Here we show how to obtain sampling variance for a slope (or a 'rate of change') when you are given measurements of a physiological trait at two points in an environmental gradient (e.g. temperature, salinity, acidity). Then, we also go on to derive sampling variance for the difference between two slopes.

### Sampling variance for a slope between two points

It may be the best to start with a real example which comparative physiologists can easily associate, namely the (temperature) acclimation response ratio (ARR). ARR can be defined as a slope for physiological responses at two different temperature points, which is:

$$
\text{ARR} = \frac{\overline{X}_{1} - \overline{X}_{2}}{T_2 - T_1},
$$

where *T* stands for temperature in Celsius and $T_2 > T_1$, and $\overline{X}_1$ and $\overline{X}_2$ are the average physiological responses at two temperature points $T_1$ and $T_2$.

<!--- Dan we may need to use bar_X rather than just X or M for consistence in the main text --->

To obtain the sampling variance for this equation (slope), we will need to know several basic properties of variance. Let's assume $X_1$ is a random variable which means that this variable has a mean and standard deviation (variance). Multiplying a constant ($a$) will change the variance by the square of that constant ($a^2$) while adding or subtracting the constant ($b$) does not change the variance of $X$. This can be summarized as:

$$
\text{Var}(aX_1 \pm b)=a^2\text{Var}(X_1)
$$

Also, when adding two random variables ($X_1$ and $X_2$), the combined variance is the sum of the variance of$X1$ and and variance of $X2$ and 2 times the covariance between $X_1$ and $X_2$. This relationship can be written as:

$$
\text{Var}(X_1 \pm X_2)= \text{Var}(X_1)+ \text{Var}(X_2) \pm 2\text{Cov}(X_1,X_2) \\
= \text{Var}(X_1)+ \text{Var}(X_2) \pm 2\text{Cor}(X_1,X_2)\sqrt{ \text{Var}(X_1))\text{Var}(X_2) },
$$ where the co-variance $\text{Cov}(X_1,X_2)$ equals to the correlation multiplied by the square-root of the two variances $2\text{Cor}(X_1,X_2)\sqrt{ \text{Var}(X_1))\text{Var}(X_2) }$.

Importantly, when $X_1$ and $X_2$ are independent of each other, the covariance is 0. More concretely, if measurements are done on two different groups of animals at two different temperature points ($T_1$ and $T_2$), then the co-variances between these two sets of measurements are 0.

Therefore, when $X_1$ and $X_2$ are indepedent, we can obtain the sampling variance for ARR as:

$$
\text{Var}(\text{AAR}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{SD^2_{1}} {N_{1}} + \frac{SD^2_{2}} {N_{2}} \right),
$$

where $SD_{1}$ and $SD_{2}$ and $N_1$ and $N_2$ are standard deviations and sampling sizes at temperature $T_1$ and $T_2$. Some may still find it difficult to see how we got this equation. Let us explain further. The (sampling) standard error for $X_1$ is $SE_1 = SD_{1} / \sqrt{N_{1}}$ -- we think many of you know this as the formula to obtain *SE* from *SD*. Given this, the sampling variance for $X_1$ is $SD^2_{1} / N_{1}$, which you see as an element of the sampling variance of ARR along with the sampling variance for $X_2$, which is $X_1$ is $SD^2_{1} / N_{1}$. the term $(1/(T_2 - T_1))^2$ comes from recognizing $1/(T_2 - T_1)$ as a constant.

Now, it is not uncommon that we use the same group of animals (or plants) at two temperature points. Then, we need to add the co-variance between $X_1$ and $X_2$ ($\text{Cov}(X_1,X_2)$); note that as above, the covariance equals the corresponding correlation multiplied by square-root of two (sampling) variances. Therefore, the sampling variance can be now written as:

$$
\text{Var}(\text{AAR}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( 
\frac{SD^2_{1}} {N_1} + \frac{SD^2_{2}} {N_2} - 2r \sqrt{\frac{SD^2_{1}} {N_1}  \frac{SD^2_{2}} {N_2}} 
\right).
$$ By assuming the number of animals (*N*) were the same at the two temperature points, we can slightly simplify this formula:

$$ 
\text{Var}(\text{AAR}) = \left( 
\frac{1} {T_2 - T_1} 
\right)
\left( 
\frac{SD^2_{1} + SD^2_{2}  - 2rSD_{1}SD_{2}} {N} 
\right).
$$

where *r* is the correlation between a set of measurement $X_1$ and $X_2$ form the same individuals at two point $T_1$ and $T_2$. As you may notice, we need raw data to calculate *r*. So in reality, we often needs to assume a certain value of *r*. Noble et al (Mol Eco; REF) discuss why we can assume $r = 0.5$ and this may be the most reasonable thing to do when we do not have any estimate of *r*.

### Sampling variance for the difference between two slopes

Now let us assume that we now want to know the difference between two different ARR values: female ARR, $\text{ARR}_{f}$ and Male ARR, $\text{ARR}_{m}$. Such a difference (ARRD) can be written as:

$$
\text{ARRD} = \text{ARR}_{f} - \text{ARR}_{m}.
$$

Using the basic properties of variance and the results from the above, the sample variance for ARRD can be derived, when measurements at two temperature points are independent, as:

$$
\text{Var}(\text{ARRD}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( \frac{SD^2_{f1}} {N_{f1}} + \frac{SD^2_{f2}} {N_{f2}} +
 \frac{SD^2_{m1}} {N_{m1}} + \frac{SD^2_{m2}} {N_{m2}}
\right),
$$ where subscripts *f* and *m* stands for 'females' and 'males'.

Similarly, the dependent version of the sampling variance can be written as:

$$
\text{Var}(\text{ARRD}) = \left( \frac{1} {T_2 - T_1} \right)^2 
\left( 
\frac{SD^2_{f1} + SD^2_{f2}  - 2rSD_{f1}SD_{f2}} {N_f}   +
\frac{SD^2_{m1} + SD^2_{m2}  - 2rSD_{m1}SD_{m2}} {N_m}
\right).
$$

Now we have derive sampling variances for 'temperature' ARR and ARRD for both independent and dependent cases. But these formulas can be applied to changes in salinity, PH, oxygen and other abotic factors. Also, the constant components of these formulas can be flexibly adjusted. For example, it may be possible that males and females were measured in slightly different temperatures; say females at $T_3$ and $T_4$ ($T_4 > T_3$) and male at $T_5$ and $T_6$ ($T_6 > T_5$) like below:

$$
\text{Var}(\text{ARRD}) = \left( \frac{1} {T_4 - T_3} \right)^2 
\left( 
\frac{SD^2_{f3} + SD^2_{f4}  - 2rSD_{f3}SD_{f4}} {N_f}   +
\right) +
\left( \frac{1} {T_6 - T_5} \right)^2 
\left( 
\frac{SD^2_{m5} + SD^2_{m6}  - 2rSD_{m5}SD_{m6}} {N_m}
\right).
$$

Also, note that the difference between slopes does not need to be that of males and females. It can be between experimental and control groups or comparing any two groups or situations.

### The delta method

It is often limited when we can use the basic properties of variance to derive sampling variance. Therefore, we introduce a practical and widely applicable method to obtain approximate variance when the basic properties of variance cannot be applied. The method is widely known as the delta method, which can be written as:

Here, we use the delta method to derive the sampling variance for the log repose ratio, which is:

$$
\text{Var}(f(X_1)) \approx \text{Var}(X_1)* \left( 
f'(X_1)
\right)^2,
$$ where $f(X_1)$ represents the function of the random variable $X_1$, and importantly, $f'(X_1)$ is the first derivative of $f(X_1)$. Probably it is hard to understand what does this really means without a concrete example. So let's derive the sampling variance for $\ln{RR}$, which is:

$$
\ln{RR} = \ln \left ( \frac{X_{1}}{X_{2}}\right ) = \ln{X_1} -\ln{X_2}.
$$ Here, $f(X_1) = \ln{X_1}$. By applying $f'(X_1) = 1/{X_1}$ to the delta method and using the variance's basic properties, we have:

$$
\text{Var}(\ln{RR}) \approx  \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right ) 
- 2r\sqrt{\frac{SD_{1}^2}{N_{1}X_{1}^2}}\sqrt{\frac{SD_{2}^2}{N_{2}X_{2}^2}},
$$

where $r$ is 0 when two sets of measurements are independent (therefore, the term including $r$ disappears), while $r$ is the correlation between $\ln{X_1}$ and $\ln{X_2}$. As discussed, we usually do not have an estimate for this correlation coefficient.

Finally, we note that by using the basic properties of variance and the delta method, one can derive sampling variance for most of cases. Indeed, these were enough for us to derive samping variance for 'new' effect sizes (e.g. $lnQ_{10}$; see the main text) and as in Nakagawa et al. (2015, MEE) and Senior et al. (2020, RSM).

### 2021-07-11 MS/ms.Rmd
boxes done - a bit of corrections for ms.Rmd too

$$
\begin{equation} 
  \begin{aligned}
  lnRR = ln\left ( \frac{X_{1}}{X_{2}}\right )
  \end{aligned}
  (\#eq:lnRR)
\end{equation} 
$$

$$
\begin{equation} 
  s_{lnRR} = \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right )
  (\#eq:slnRR)
\end{equation} 
$$ In equation \@ref(eq:lnRR), $X_{1}$ is the mean of group 1, often a control group, where as $X_{2}$ is the mean of group 2 (e.g., a treatment group). The mean of $X_{i}$ for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the measurement variable is ratio scale. Log transformation of this ratio makes the effect size normally distributed (most commonly assumed by meta-analytic models). Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where, $SD_{1}^2$ and $SD_{2}^2$ is the sample variance in group 1 and 2, respectively and $N_{1}$ and $N_{2}$ is the sample size in group 1 and 2.

@@

$$
\begin{equation} 
  s_{lnRR_{Q_{10}}} = \left( \frac{SD_{2}^2}{R^2_{2}N_{2}} + \frac{SD_{1}^2}{R^2_{1}N_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
  (\#eq:Vlnq10)
\end{equation} 
$$

@@

$$
\begin{equation} 
lnVR_{Q_{10}} = ln\left( \frac{SD_{2}}{SD_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10VR)
\end{equation} 
$$ $$
\begin{equation} 
s_{lnVR_{Q_{10}}} = \left( \frac{1}{2\left( N_{2}-1 \right)} + \frac{1}{2\left(N_{1} - 1\right)} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
(\#eq:slnq10VR)
\end{equation} 
$$

@@

$$
\begin{equation} 
lnCVR_{Q_{10}} = ln\left( \frac{\text{CV}_{2}}{\text{CV}_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10CVR)
\end{equation} 
$$ $$
\begin{equation} 
s_{lnCVR_{Q_{10}}} = \left[ \frac{(\text{SD}_{1})^2}{N_{1}({R}_{1})^2} + \frac{(\text{SD}_{2})^2}{N_{2} ({R}_{2})^2} + \frac{1}{2\left( N_{1}-1 \right)} + \frac{1}{2\left(N_{2} - 1\right)} \right]{ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
(\#eq:slnq10CVR)
\end{equation} 
$$

>>>>

$$
\begin{equation} 
  \begin{aligned}
  \ln{RR} = \ln\left ( \frac{X_{1}}{X_{2}}\right )
  \end{aligned}
  (\#eq:lnRR)
\end{equation} 
$$

$$
\begin{equation} 
  s^2_{lnRR} = \left ( \frac{SD_{1}^2}{N_{1}X_{1}^2}\right ) + \left ( \frac{SD_{2}^2}{N_{2}X_{2}^2}\right )
  (\#eq:slnRR)
\end{equation} 
$$ In equation \@ref(eq:lnRR), $X_{1}$ is the mean of group 1, often a control group, where as $X_{2}$ is the mean of group 2 (e.g., a treatment group). The mean of $X_{i}$ for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the measurement variable is ratio scale. Log transformation of this ratio makes the effect size normally distributed (most commonly assumed by meta-analytic models). Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where, $SD_{1}^2$ and $SD_{2}^2$ is the sample variance in group 1 and 2, respectively and $N_{1}$ and $N_{2}$ is the sample size in group 1 and 2.

@@

$$
\begin{equation} 
  s^2_{lnRR_{Q_{10}}} = \left( \frac{SD_{2}^2}{R^2_{2}N_{2}} + \frac{SD_{1}^2}{R^2_{1}N_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
  (\#eq:Vlnq10)
\end{equation} 
$$

@@

$$
\begin{equation} 
lnVR_{Q_{10}} = ln\left( \frac{SD_{2}}{SD_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10VR)
\end{equation} 
$$ $$
\begin{equation} 
s^2_{lnVR_{Q_{10}}} = \left( \frac{1}{2\left( N_{2}-1 \right)} + \frac{1}{2\left(N_{1} - 1\right)} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
(\#eq:slnq10VR)
\end{equation} 
$$

@@

$$
\begin{equation} 
lnCVR_{Q_{10}} = ln\left( \frac{\text{CV}_{2}}{\text{CV}_{1}} \right){ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }
(\#eq:lnq10CVR)
\end{equation} 
$$ $$
\begin{equation} 
s^2_{lnCVR_{Q_{10}}} = \left[ \frac{(\text{SD}_{1})^2}{N_{1}({R}_{1})^2} + \frac{(\text{SD}_{2})^2}{N_{2} ({R}_{2})^2} + \frac{1}{2\left( N_{1}-1 \right)} + \frac{1}{2\left(N_{2} - 1\right)} \right]{ \left(\ \frac{10^{\circ}C}{T_{2}-T_{1}} \right) }^2
(\#eq:slnq10CVR)
\end{equation} 
$$
